Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-01 linear_adia.tex
edits to 5, mainly 5.3.  Please check, especially the footnote about the correspondence with appendix C which maybe isn't exact?

In \S\ref{relax_pert} we show that even an infinitesimal increase in the cooling time,
starting from $\beta = 0$, slows the growth rate of the VSI.  In \S\ref{disp_relax} we 
derive exact solutions to the simplified VSI model developed above (Eq.\ \ref{nearly_iso_explicit}).
In \S\ref{iso_vsi_beta_crit} we reach our main result, the maximum cooling time above which 
VSI growth is suppressed.  We heuristically argued for this result in
\S\ref{vsi_require}, and will numerically
verify it in \S\ref{numerical}.

@@

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth the VSI.  In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave frequency,
\hat{\omega}_c$  at marginal stability to the VSI

%Our starting point is Eq. \ref{relax_cond1}---\ref{relax_cond2} which
%approximates the dispersion relation Eq. \ref{relax_disp} assuming marginal
%stability and $\khat^2\gg 1$. We consider $M=0$ (the 
%fundamental mode) or  $M=O(1)$ and low frequency $|\hat{\omega}|\ll
%1$ such that $|\hat{\omega}\khat|$ is $O(1)$. We 
%also assume $|q|$ and $\beta$ are $O(1)$ or smaller. 
%
%For a thin disk
%($h \ll 1$) we have 
\begin{subequations}\begin{align}
 \left(\hat{\omega}\khat\right)^4  = &  \left[1 +
    \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}\khat\right)^2 \label{relax_cond_simp1}\\ 
 &+ M(M+1)\left(1-h^2q^2\khat^2\right)
    , \notag \\
   \left(\hat{\omega}\khat\right)^2 = & \frac{\beta}{h q} (\gamma-1)(1+2M)^2
   \left(\hat{\omega}\khat\right) \label{relax_cond_simp2}\\
   $- 2M(M+1) , \notag
\end{align}\end{subequations}
These equations hold for $\beta_c \ll 1$ and $\hat{k} \sim 1/|\

Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2} is a pair of
simultaneous equations for $\hat{\omega}$ and $\beta$ \emph{assuming} marginal 
stability. %  (It is important to note that marginal stability is an
% assumption.)
If $\beta\ll 1$ and the disk is sufficiently thin so that  
$|h q \khat |^2\lesssim 1$,

An approximate solution to Eq. \ref{relax_cond_simp1} for the frequency
\begin{align}
  \left(\hat{\omega}\khat\right)^2 \simeq (M+1)\left(1 -
    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
holds for either (a) $M= 0$ and arbitrary $|h q \hat{k}|$ or (b) $|h q \hat{k}| \ll 1$ and arbirtary $M \sim O(1)$. 
Note that for $q=0$, Eq. \ref{marg_freq_approx} reduces to 
the dispersion relation for low-frequency inertial waves in vertically
isothermal disks without vertical shear (see Appendix
\ref{stable_novshear}). Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta &= \frac{h|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}\left[1 + \frac{M(2M-1)|h q
      \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
Since $|h q \khat|^2\lesssim 1$ by assumption, we have  
\begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1}, 
\end{align}
where $\beta_\mathrm{crit}$ is the critical thermal timescale for
the fundamental mode ($M=0$). 
We comment the expression for $\beta_\mathrm{crit}$ may in fact be
obtained without assuming it is much less than unity from the onset or
$|h q\khat|\lesssim 1$. % This is because for $M=0$, 
% Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} simplify 
We will  
confirm $\beta_\mathrm{crit}$ numerically in \S\ref{bcrit_num_test}.

Eq. \ref{iso_vsi_cond} suggests that as the thermal timescale $\beta$
is increased from zero, higher order modes with $M>0$
will be stabilized before the fundamental mode. We indeed observe this
trend in numerical calculations (\S\ref{therm_relax_eff}), despite
adopting somewhat restrictive assumptions for $M>0$ in order to arrive
at the simple expression Eq. \ref{bcrit_gen}.

%Rapid thermal relaxation is required to overcome a stabilizing
%vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. 
%Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
%global disturbances (which have small $M$).

We can interpret  $\beta_\mathrm{crit}$ as the thermal timescale below
which buoyancy becomes ineffective and perturbations are effectively
isothermal so the VSI can operate.  We see that in a thin disk, a very short thermal timescale
is required ($\beta_\mathrm{crit}\ll 1$) because
$\beta_\mathrm{crit}\propto h$. Physically, this is because
vertical shear is weak in a thin disk, so even a small entropy
gradient can significantly reduce VSI growth rates.

% Here we estimate the maximum thermal relaxation timescale
% $\beta$ that allows the fundamental VSI mode for $\khat\gg1$.

% At marginal stability, the eigenfrequency $\hat{\upsilon}=\hat{\omega}$
% is real. The dispersion relation for the fundamental mode is then
% \begin{align}\label{relax_disp_fund}
%   0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
% \end{align} 
% Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} and considering 
% $\khat\gg1$, we find
% \begin{align}
%   0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2h q
%   \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
%   &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
%   0=& \beta(1-\gamma)\khat^2 + h q \khat^2 (\hat{\omega}\khat)
%   - 2\beta (\hat{\omega}\khat)^2 - h q \gamma^2\beta^2 (\hat{\omega}\khat)^3 \notag\\
%     &+ 2\beta\gamma(\hat{\omega}\khat)^4.  
% \label{relax_cond2}
% \end{align}
% This is a pair of simultaneous equations for
% $\hat{\omega}$ and $\beta$. Now, low-frequency axisymmetric waves in
% the absence of vertical shear are stable and  satisfy
% $\hat{\omega}\propto \khat^{-1}$ (Appendix \ref
% {stable_novshear}, Eq. \ref{iso_disp_full}).  We are thus motivated to
% seek solutions to Eq. \ref{relax_cond1}---\ref{relax_cond2} with  
% % we wish to seek a criterion independent of k 
% % \begin{align}
% %   \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
% %\end{align}
% \begin{align}
%   \khat \gg1, \quad |\hat{\omega}|\ll 1 ,\quad
%   |\hat{\omega}\khat|=O(1). 
% \end{align}
% Then
% \begin{align}
%   \hat{\omega}\khat = \frac{(\gamma-1)\beta}{h q}
% \end{align}
% from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
% that
% \begin{align}\label{beta_crit0}
%   \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(h
%         q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
% \end{align}
% at marginal stability for perturbations with large $\khat$.

% From our discussion in \S\ref{relax_pert}, which indicate introducing
% a finite thermal relaxation is stabilizing, we expect
% Eq. \ref{beta_crit0} is an upper limit. For a thin  
% disk $h\ll 1$, we thus infer 
% \begin{align}\label{iso_vsi_cond}
%   \beta \lesssim \frac{|h q|}{\gamma-1} \equiv
%   \beta_\mathrm{crit} 
% \end{align}
% is required for the fundamental VSI to operate. That is, when
% $\beta<\beta_\mathrm{crit}$, perturbations are  effectively isothermal. 
% We will confirm Eq. \ref{iso_vsi_cond} numerically in
% \S\ref{bcrit_num_test}.

>>>>

In \S\ref{relax_pert} we show that even an infinitesimal increase in the cooling time,
starting from $\beta = 0$, slows the growth rate of the VSI.  In \S\ref{disp_relax} we 
derive exact solutions to the simplified VSI model developed above (Eq.\ \ref{nearly_iso_explicit}).
In \S\ref{iso_vsi_beta_crit} we reach our main result, the maximum cooling time above which 
VSI growth is suppressed.

@@

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth the VSI.  In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave frequency,
$\hat{\omega}_c$  at marginal stability to the VSI:
\begin{subequations}\begin{align}
 \left(\hat{\omega}_c\khat\right)^4  = &  
% \left[1 + \gamma\beta^2\left(1-\gamma\right)(1+2M)^2\right]\left(\hat{\omega}_c\khat\right)^2 \label{relax_cond_simp1}\\ 
 \left(\hat{\omega}_c\khat\right)^2  + M(M+1)\left(1-h^2q^2\khat^2\right) , \label{relax_cond_simp1} \\
   \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta}{h q} (\gamma-1)(1+2M)^2
   \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
   &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations hold for $\beta_c \ll 1$ and $\khat \sim 1/\hat{\omega}_c \gg1$,
as explained in Appendix \ref{disp_neut_limit}.

Eq. \ref{relax_cond_simp1} gives an approximate result for the frequency
\begin{align}
 \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left[1  + O( M|h q \khat|^2)\right]  \label{marg_freq_approx}
%  \left(\hat{\omega}_c\khat\right)^2 \simeq (M+1)\left(1 -
%    \frac{M|h q \khat|^2}{2M+1}\right).  \label{marg_freq_approx}
\end{align}
which holds for either (a) $M= 0$ and arbitrary $|h q \hat{k}|$ or (b) $|h q \hat{k}| \ll 1$ and arbirtary $M \sim O(1)$. 
These distinct conditions mean that our results are most robust for the fundamental vertical mode ($M = 0$) but also
hold for higher order vertical modes when radial wavelengths are not too short.
Note that Eq. \ref{marg_freq_approx} 
reduces (\emph{nearly}?)\footnote{\emph The correspondence is only exact 
if $L = M+1$.  Is this the case (in which case at least note in the appendix) or is there a discrepancy (in which case 
we can't say it reduces)?}
 to the dispersion relation for low-frequency inertial waves 
%in vertically isothermal disks without vertical shear 
(see Appendix\ref{stable_novshear} for details).

Inserting Eq. \ref{marg_freq_approx} into
Eq. \ref{relax_cond_simp2}, we obtain 
\begin{align}\label{bcrit_gen}
  \beta_c &\simeq  \frac{h|q|}{(\gamma - 1)}
  \frac{\sqrt{1+M}}{(1+2M)}
%  \left[1 + \frac{M(2M-1)|h q   \khat|^2}{2(2M+1)^2}\right].
%  &\lesssim \beta_\mathrm{crit}\notag
\end{align}
where the above-mentioned approximations still hold, but the correction term ($O(M |h q \khat|^2)$) is now suppressed.
Eq. \ref{bcrit_gen} says that higher $M$ modes have lower $\beta_c$, i.e.\ require faster cooling for VSI growth.
This trend is confirmed numerically in \S\ref{therm_relax_eff} (and the trend holds even when the restrictive assumption
on $|h q \hat{k}|$ is dropped).

We are most interested in the longest cooling time which allows growth for any $M$.  
This criterion is clearly set by the fundamental, $M = 0$, mode.  Our cooling criterion for 
VSI growth is thus 
 \begin{align}\label{iso_vsi_cond}
  \beta\leq \beta_\mathrm{crit}  \equiv
  \frac{h|q|}{\gamma-1} . 
\end{align}
The thin disk approximation, $h \ll 1$, justifies the $\beta_c \ll 1$ approximation.
 We heuristically explain $\beta_\mathrm{crit}$ in
\S\ref{vsi_require}, and numerically test its validity in \S\ref{bcrit_num_test}.

%Eq. \ref{iso_vsi_cond} suggests that as the thermal timescale $\beta$
%is increased from zero, higher order modes with $M>0$
%will be stabilized before the fundamental mode. We indeed observe this
%trend in numerical calculations (\S\ref{therm_relax_eff}), despite
%adopting somewhat restrictive assumptions for $M>0$ in order to arrive
%at the simple expression Eq. \ref{bcrit_gen}.

%We can interpret  $\beta_\mathrm{crit}$ as the thermal timescale below
%which buoyancy becomes ineffective and perturbations are effectively
%isothermal so the VSI can operate.  We see that in a thin disk, a very short thermal timescale
%is required ($\beta_\mathrm{crit}\ll 1$) because
%$\beta_\mathrm{crit}\propto h$. Physically, this is because
%vertical shear is weak in a thin disk, so even a small entropy
%gradient can significantly reduce VSI growth rates.
